---
title: "Extracting climate data for monkeyflower populations"
author: "Kelly Carscadden"
date: "18/01/2021"
output: html_document
---

Aim: given coordinates of study populations, extract data for several hydrological variables across 50 years, from California-wide climate rasters

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gdalUtils)
library(sf)
library(raster)
```

```{r}
#load dataset containing coordinates of study populations
exp_coord_df <- read_csv("exp_coord_df.csv")
```

Make spatial points dataframe and re-project it to match the climate rasters
```{r}
#set coordinates, make into a spatial points df
coordinates(exp_coord_df)<- c("long", "lat") 

#assign the projection system
proj4string(exp_coord_df) <- CRS("+proj=longlat +datum=WGS84") 

#transform lat/lon from sites into the coordinate system used in the climate rasters
popTrans<-spTransform(exp_coord_df
  ,CRS("+proj=aea +lat_1=34 +lat_2=40.5 +lat_0=0 +lon_0=-120 +x_0=0 +y_0=-4000000 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs"))

#make into simple features object for ease of working with below
popTrans <- popTrans %>% as('sf')
```

Specify the hydrological variables and years of interest. Hydrological variable codes follow http://climate.calcommons.org/dataset/CA-BCM-2014-monthly. Data from 2011-2014 stored as ascii files labeled by month
```{r}
#e.g., aet is the code for actual evapotranspiration
vars <- c("tmx", "tmn", "run", "rch", "ppt", "pet", "pck", "cwd", "aet")
years <- 2011:2014
months <- c("jan","feb","mar","apr","may","jun","jul","aug","sep","oct","nov","dec")
```

For each hydrological variable and year, iterate across and stack monthly raster files, and extract data at each population location

- script adapted from Prahlada Papper (UC Berkeley) and Mike Koontz (CU Boulder)
```{r}
for(i in seq_along (vars)){
  for(j in seq_along(years)){
   this_var <- vars[i]
    this_year <- years[j]
   
    #start with january
     clim <- raster(paste0("/", this_var, this_year, "jan.asc"))
     crs(clim) <- st_crs(3310)$proj4string
       #'+proj=aea +datum=NAD83 +lat_1=34 +lat_2=40.5 +lat_0=0 +lon_0=-120 +x_0=0 +y_0=-4000000'

    #create raster stack by adding additional months
    for(m in months[2:12]) { 
      clim <- stack(clim, raster(paste0(this_var, this_year, m, ".asc"))) 
    }
     
    #create a directory for the output, if it does not already exist 
    if(!dir.exists("Carscadden")) {
      dir.create("Carscadden", recursive = TRUE)
    }

    #extract climate data at population coordinates. Export data
    ddpop <-extract(normal, popTrans)
    saveRDS(ddpop, file = paste0("Carscadden/", "CA_BCM_HST_Monthly_", this_var, "_",
                                 this_year, "_pops.rds"))
  }
}
```