---
title: "Bioclim and PCA"
author: "Kelly Carscadden"
date: "04/03/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dismo) #for biovars funct
library(RcppRoll) #for sliding window
library(GGally) #for correlation plot
library(vegan) #for PCA
library(viridis) #for colors
```

Aggregate climate data so is meaningful for the growing season hydrology of focal populations.

1. Using a sliding 3-month-window analysis similar to Bioclim, determine min/max quarterly climate estimates (e.g., max temperature of warmest quarter)

```{r}
#specify years and variables
years <- 1964:2014
vars <- c("tmx", "tmn", "ppt", "pet", "pck", "cwd", "aet", "run", "rch") 

#for each year, summarize the climate variables
for (i in seq_along(years)){
   this_year <- years[i] 
   for (j in seq_along(vars)) {
          this_var <- vars[j] #need to summarize each variable
          
          #read in raster for each variable
          assign(paste0(this_var, "_", this_year, "_dat"),
                 readRDS(paste0("data/extract/", "CA_BCM_HST_Monthly_", this_var, "_",
                                 this_year, "_pops.rds")))
    }
  
    #create output directory if it does not exist
    if(!dir.exists("data/clim_summary")) {
      dir.create("data/clim_summary", recursive = TRUE)
    }
    
    #need ppt, tmn, tmx for biovars, for each year (e.g., calculates max ppt in hottest quarter)
    #in output, rows = bioclim vars, cols = sites
    #generates 19 bioclim variables
    clim_agg <- as.data.frame(biovars(prec = get(paste0("ppt_", this_year, "_dat")),
            tmin = get(paste0("tmn_", this_year, "_dat")),
            tmax = get(paste0("tmx_", this_year, "_dat")))
    )
    
    #aggregate other clim variables, in 3-mo sliding windows
    others <- vars[!vars %in% c("ppt", "tmn", "tmx")]

    for(k in seq_along(others)){
      other_var <- others[k]
      label <- paste0("max_", other_var)
      
      clim_agg <- cbind(clim_agg, 
                                   apply(X = get(paste(other_var, this_year, "dat", sep = "_")),
                                        MARGIN = 1, #across cols, by row
                                        #find max 3-mo clim value. Window ID'd by start month
                                        FUN = roll_mean,
                                        n = 3, align = "left", na.rm = F , partial = F) %>%
                                        apply(MARGIN = 2, FUN = max) 
                                     )
      #fix name of new variable (there are 19 bioclim variables)
      colnames(clim_agg)[19+k] <- label
    }
    
    #export
    saveRDS(clim_agg, file = paste0("data/clim_summary/", "CA_BCM_HST_Monthly_",
                                 this_year, "_summary.rds"))
    
}

```


2. thin climate variables based on collinearity and relevance
- have made summaries 1964 - 2014 - using this fixed interval for all pops

```{r}
#make big data stack of all summaries, sites
##melt so year is a column (not included in PCA)
##add in spPop labels for subsetting - combine abbrev_sp and code in exp_coord_df

summary_df <- NULL
years <- c(1964:2014)

for(i in seq_along(years)){
  this_year <- years[i]
  yr_df <- readRDS(paste0("data/clim_summary/CA_BCM_HST_Monthly_", this_year, "_summary.rds"))
  
  #annotate yrs - can add site labels after
  yr_df$yr <- this_year
  
  #join to other yrs
  summary_df <- rbind(summary_df, yr_df)
}

#add site labels
rep_yrs <- length(years) #51 yrs

#need non-spatial object to pull labels
exp_coord_df <- read_csv("exp_coord_df.csv")
exp_coord_df <- exp_coord_df %>% tidyr::unite(col = "sp_pop", c(abbrev_sp, code), remove = F) 

summary_df$sp_pop <- rep(exp_coord_df$sp_pop, rep_yrs)
```

```{r}
#weed out bioclim vars like annual mean temp, isothermality, seasonality, other measures of var
summary_df_pared <- summary_df %>% dplyr::select(-bio1, -bio2, -bio3, -bio4, -bio7,
                                                    -bio12, -bio15)
#explore correlations & variation in diff variables
GGally::ggpairs(summary_df_pared[1:18]) #omit yr, sp_pop
#throwing in all data for all sites, years, for exploratory viz

#several variables (eg. temp of warmest quarter and temp of driest quarter) ~ collinear given mediterranean climate - can pare down

summary_df_pared <- summary_df_pared %>% dplyr::select(c(5, 6, 9, 10, 13:20))
#tidy up labels
summary_clim_pared <- summary_df_pared %>% rename("max_tmp" = bio10, 
       "min_tmp" = bio11,
       "max_ppt" = bio16,
       "min_ppt" = bio17)#bio 10 aka mean temp of warmest 1/4, bio17 = mean ppt of driest 1/4
rm(summary_df)

#export
#write.csv(summary_clim_pared, "summary_clim_pared.csv", row.names = F)
```


```{r}
#save correlation plot of remaining clim variables
pdf("mim clim ggpairs all yrs and sites.pdf",
    width = 16,
    height = 9)

  GGally::ggpairs(summary_clim_pared[1:10])
dev.off()
```


3. Use PCA to further reduce dimensionality

```{r}
clim.pca <- rda(decostand(summary_clim_pared[,c(1:10)], method = "standardize")) 

#determine %Var explained by each PCA axis
pca_var <- summary(clim.pca)
round(pca_var$species, 2)
```

Each species contains multiple populations. Visualize the climatic variation in each species and populations' evolutionary history by plotting polygons from the climatic PCA

```{r}
#plot and calc avg pop ellipse area, for each sp
summary_clim_pared <- summary_clim_pared %>% 
  separate(sp_pop, into = c("sp", "pop"), sep = "_", remove = F) %>%
  mutate(pop = as.factor(pop),
         sp = as.factor(sp))

site_scores <- scores(clim.pca)$sites
```

```{r}
#join pril and prip (subspecies that belong to one species)
summary_clim_pared_filled <- summary_clim_pared_filled %>%
  dplyr::mutate(sp = as.character(sp))

summary_clim_pared_filled$sp[summary_clim_pared_filled$sp == "PRIP"] <- "PRIM"
summary_clim_pared_filled$sp[summary_clim_pared_filled$sp == "PRIL"] <- "PRIM"

#make a plotting function that will loop across species
sp_list <- unique(summary_clim_pared_filled$sp)
sp_cols <- viridis(length(sp_list)) #one color per sp

areas_df <- NULL
perimeters_df <- NULL

for(i in seq_along(sp_list)){
  #subset data for sp
  this_sp <- sp_list[i]
  this_col <- sp_cols[i]
  
  sub <- summary_clim_pared_filled %>% filter(sp == this_sp)
  pops <- unique(sub$sp_pop)
  ind <- which(summary_clim_pared_filled$sp == this_sp)
  scores <- site_scores[ind,]
  
  #plot polygons for the species as a whole, and its populations
  plot(clim.pca, display = "sites", scaling = "symmetric", type = "n", 
            ylim = c(-1, 0.8))
  points(scores, pch = 21, bg = "grey66", col = "black")
  sp_hull <- with(summary_clim_pared_filled, ordihull(clim.pca, sp, draw="lines",
                                     col = this_col, show.groups = this_sp, lwd = 2))
  pop_hull <- with(summary_clim_pared_filled, ordihull(clim.pca, sp_pop, draw="polygon",
                                     col = this_col, show.groups = pops,
                                     label = F, cex = 0.5))
  
  #save perimeters of polygons, for calculating overlap
  temp <- data.frame(scores(pop_hull),
                     sp_pop = attributes(scores(pop_hull))$hulls)
  
  perimeters_df <- rbind(perimeters_df, temp)
  
  #save polygon area calculations (estimate of climate breadth) and centroids (estimate of climatic niche position)
  sp_df <- summary(sp_hull)
  pop_df <- summary(pop_hull)
  
  dat <- data.frame(sp = this_sp, sp_pop = c("whole sp", colnames(pop_df)),
             PC1_centroid = c(sp_df[1], pop_df[1,]),
             PC2_centroid = c(sp_df[2], pop_df[2,]),
             area = c(sp_df[3], pop_df[3,]))
  
  areas_df <- rbind(areas_df, dat)

}#end of outer for

#export
#write.csv(areas_df, "clim_polygon_areas.csv", row.names = F)

#export
#write.csv(perimeters_df, "clim_polygon_perimeters.csv", row.names = F)
```

Overlay all species' climate polygons in a single plot, for ease of comparison
```{r}
plot(clim.pca, display = "sites", scaling = "symmetric", type = "n", 
            ylim = c(-1, 0.8)) #updated in V2 to 0.8
  #points(site_scores, pch = 21, bg = "grey66", col = "black")
  sp_hull <- with(summary_clim_pared_filled, ordihull(clim.pca, sp, draw="polygon",
                                     col = sp_cols))

#re-ordering species to increase visibility in plot
sp_rev <- summary_clim_pared_filled %>% 
  mutate(sp = fct_relevel(sp, "PRIM", "MOSC", "CARD", "BICO", "KELL", "GLAU", "DOUG")) 
cols_rev <- c("#FDE725FF", "#8FD744FF", "#443A83FF", "#440154FF", "#35B779FF",
              "#21908CFF", "#31688EFF")    

#save plot
pdf("species climate polygons V2.pdf",
      width = 4,
      height = 4)

plot(clim.pca, display = c("sites"), scaling = "symmetric", type = "n", 
            ylim = c(-1, 0.8),
     xlab = "PC1 (50.43%)", #UPDATE THESE IF CHANGING THE PCA
     ylab = "PC2 (15.04%)")
  sp_hull <- with(sp_rev, ordihull(clim.pca, sp, draw="polygon",
                                     col = cols_rev))

dev.off()    
```